---
title: "Pitching the Probabilities Code"
output: html_notebook
---

By: Amanda Poor

Below is the necessary code I used to create the output for the probabilities, statistic test, and visual. Please reference the formal documentation of my report found on my personal website https://amandapoor.net

```{r}
#load packages 
library(dplyr)
library(tidyr)
library(EnvStats)
```

```{r}
#below is the dataset that has been created through merging each MLB's team pitch data from the 2021 regular season only including innings 1-9. 

#save the regseason.csv to your computer and then uncomment line 17 to select the dataset. 
#regseason<-read.csv(file.choose())

regseason
```

```{r}
#package 'dplyr' used to mutate the dataframe
# creation of a 'count' variable, combining the existing balls and strikes variables

regseason<-regseason %>%
  mutate(regseason, 
         count = ifelse(regseason$balls=="3" & regseason$strikes=="2","3_2",
                    ifelse(regseason$balls=="3" & regseason$strikes=="1","3_1",
                        ifelse(regseason$balls=="3" & regseason$strikes=="0","3_0",
                            ifelse(regseason$balls=="2" & regseason$strikes=="2","2_2",
                                ifelse(regseason$balls=="2" & regseason$strikes=="1","2_1",
                                    ifelse(regseason$balls=="2" & regseason$strikes=="0","2_0",
                                        ifelse(regseason$balls=="1" & regseason$strikes=="2","1_2",
                                            ifelse(regseason$balls=="1" & regseason$strikes=="1","1_1",
                                                ifelse(regseason$balls=="1" & regseason$strikes=="0","1_0",
                                                    ifelse(regseason$balls=="0" & regseason$strikes=="2","0_2",
                                                        ifelse(regseason$balls=="0" & regseason$strikes=="1","0_1",
                                                            ifelse(regseason$balls=="0" & regseason$strikes=="0","0_0", NA)))))))))))))   %>%
# creation of the variable 'PBhand' (pitcher/batter handedness; bats R/L, pitches R/L)  
  mutate(regseason, PBhand = ifelse(regseason$stand=="R" & regseason$p_throws=="R","bRpR",
                                                ifelse(regseason$stand=="R" & regseason$p_throws=="L","bRpL",
                                                       ifelse(regseason$stand=="L" & regseason$p_throws=="R","bLpR",
                                                            ifelse(regseason$stand=="L" & regseason$p_throws=="L","bLpL", NA)))))
 
```

```{r}
#creation of the 'pitch_cat' variable which categorizes the pitch types into MLB's pitch categories (fastball, breaking, offspeed)

regseason<- regseason %>% mutate(pitch_cat = 
                                  ifelse(regseason$pitch_type == "FC" | regseason$pitch_type == "FF" | regseason$pitch_type == "FA"
                                   | regseason$pitch_type == "SI", "fastball", 
                                        ifelse(regseason$pitch_type == "CH" | regseason$pitch_type == "KN" | 
                                         regseason$pitch_type == "FS", "offspeed", 
                                              ifelse(regseason$pitch_type == "KC" | regseason$pitch_type == "SL"
                                               | regseason$pitch_type == "CU" | regseason$pitch_type == "CS", "breaking", NA))))

```


```{r}
# selection of the top pitch outcome descriptions; outcomes which had very few instances were dropped

reg.description<- regseason[regseason$description %in% c("ball", "foul", "hit_into_play", "called_strike", "swinging_strike", "foul_tip"), ]
```

```{r}
# creation of the 'p_outcome' variable, which separates the pitch descriptions selected above into either a desirable or undesirable outcome from the pitchers perspective. 

reg.outcome<-reg.description %>%
  mutate(p_outcome = ifelse(reg.description$description == "hit_into_play" & (reg.description$events == "single" | reg.description$events == "double"
                        | reg.description$events == "triple" | reg.description$events == "home_run"), "undesirable", 
                      
                            ifelse(reg.description$description == "hit_into_play" & 
                                (reg.description$events == "force_out" | reg.description$events == "field_out") , "desirable", 
                              
                                  ifelse(reg.description$description == "foul" | reg.description$description == "foul_tip", "desirable",
                                           
                                        ifelse(reg.description$description == "called_strike" | 
                                            reg.description$description == "swinging_strike", "desirable", 
                                              
                                               ifelse(reg.description$description == "ball", "undesirable", NA))))))

```
```{r}
# converts the following variables into factor variables

reg.outcome$p_outcome <- as.factor(reg.outcome$p_outcome)
reg.outcome$PBhand <- as.factor(reg.outcome$PBhand)
reg.outcome$pitch_type <- as.factor(reg.outcome$pitch_cat)
reg.outcome$count <- as.factor(reg.outcome$count)

```

```{r}

#subset dataframe of regseason containing varialbes of interest

reg.outcome2=subset(reg.outcome, select=c("pitch_type", "pitch_cat", "stand", "p_throws", "PBhand","balls", "strikes", "count", "description","p_outcome"))

#drop na from selected variables

reg.outcome<-reg.outcome2 %>% drop_na("PBhand", "p_outcome", "count", "pitch_cat")
```




____________________________________________________________________
____________________________________________________________________

Creation of Proportions

- Proportions were made as they are necessary for computation of Probabilities


```{r}
#total number of right handed pitchers at each count

total_RHP<-reg.outcome %>% 
  filter(PBhand == "bRpR" | PBhand =="bLpR" )%>%
  drop_na(pitch_cat, PBhand, count, p_outcome) %>%
  group_by(count) %>% 
  summarize(Total_RHP = n()) %>%
  group_by(count)
total_RHP
```
```{r}
#total number of desirable right handed pitches at each count

total_D_RHP<-reg.outcome %>% 
  filter(p_outcome== "desirable" & (PBhand == "bRpR" | PBhand =="bLpR") )%>%
  drop_na(pitch_cat, PBhand, count, p_outcome) %>%
  group_by(count) %>% 
  summarize(Total_D_RHP = n()) %>%
  group_by(count) 
total_D_RHP
```

```{r}
#combine total RHP and total desirable RHP, factored by count
D_RHP_Prop<- merge(total_RHP, total_D_RHP, by = c("count")) 

#creation of proportions for desirable outcomes by RHP
D_RHP_Prop$D_RHP_Prop <- D_RHP_Prop$Total_D_RHP/D_RHP_Prop$Total_RHP
D_RHP_Prop
```

```{r}
#desirable right handed pitches identity matrix

#take 'D_RHP_Prop' variable above and convert to matrix
D_RHP_Prop_matrix <- data.matrix(D_RHP_Prop$D_RHP_Prop)
#transpose matrix and convert to vector
D_RHP_Prop_T_matrix<-t(D_RHP_Prop_matrix)
D_RHP_Prop_vector<-as.vector(D_RHP_Prop_T_matrix)

#'D_RHP_Prop_vector' has to be a vector; 12x12 identity matrix; 'D_RHP_Prop' proportions along diagonal of identity matrix
D_RHP_Prop_I_matrix<-matrix(diag(D_RHP_Prop_vector), ncol=12)
D_RHP_Prop_I_matrix
```



```{r}
#total number of undesirable right handed pitches 

total_UD_RHP<-reg.outcome %>% 
  filter(p_outcome== "undesirable" & (PBhand == "bRpR" | PBhand =="bLpR") )%>%
  drop_na(pitch_cat, PBhand, count, p_outcome) %>%
  group_by(count) %>% 
  summarize(Total_UD_RHP = n()) %>%
  group_by(count) 
```


```{r}
#same code as above but for undesirable outcomes

UD_RHP_Prop <- merge(total_RHP, total_UD_RHP, by = c("count")) 
UD_RHP_Prop$UD_RHP_Prop <- UD_RHP_Prop$Total_UD_RHP/UD_RHP_Prop$Total_RHP
UD_RHP_Prop
```


```{r}
#undesirable right handed pitches identity matrix

UD_RHP_Prop_matrix <- data.matrix(UD_RHP_Prop$UD_RHP_Prop)
UD_RHP_Prop_T_matrix<-t(UD_RHP_Prop_matrix)
UD_RHP_Prop_vector<-as.vector(UD_RHP_Prop_T_matrix)

# 12x12 identity matrix made using the undesirable RHP proportions from UD_RHP_Prop dataframe

UD_RHP_Prop_I_matrix<-matrix(diag(UD_RHP_Prop_vector), ncol=12)
UD_RHP_Prop_I_matrix
```


Notice how both the desirable and undesirable proportions for each count add to 100%. This makes sense as they are proportions for the pitch outcomes

I will now show the code for the bRpR 'PBhand' instances. It is the same for each of the other 'PBhand' factors

____________________________________

Bats Right Pitches Right

The following code demonstrates how to create the 3x12 matrices for bRpR. After the 3x12 matrix will be multiplied by the identity matrices above and then will be made into probabilities of a desirable outcome occurring for the bRpR instance


```{r}
reg.desirable <- reg.outcome[reg.outcome$p_outcome == "desirable" , ] 

#select bRpR instance from desirable outcomes
reg.des.Pitch.bRpR <- reg.desirable[reg.desirable$PBhand == "bRpR", ]   

#create table pitch_cat by count
reg.des.Pitch.bRpR.count<-table(reg.des.Pitch.bRpR$pitch_cat, reg.des.Pitch.bRpR$count)

#round the values to 2 decimal places
reg.des.Pitch.bRpR.count.prop<-round(prop.table(reg.des.Pitch.bRpR.count,2)*100,digits=2)
reg.des.Pitch.bRpR.count.prop
```



```{r}
#proportions of count distributions

#convert proportions to matrix
reg.des.Pitch.bRpR.count.matrix<-as.matrix(reg.des.Pitch.bRpR.count.prop)

#multiply 3x12 bRpR matrix by desirable RHP identity matrix
reg.des.pitch.bRpR.count.prop<-reg.des.Pitch.bRpR.count.matrix %*% D_RHP_Prop_I_matrix

#add column names (count)
colnames(reg.des.pitch.bRpR.count.prop) <- c('0-0', '0-1', '0-2', '1-0', '1-1', '1-2', '2-0', '2-1', '2-2', '3-0', '3-1', '3-2')
reg.des.pitch.bRpR.count.prop
```

```{r}
# same done for undesirable bRpR

reg.undesirable <- reg.outcome[reg.outcome$p_outcome == "undesirable", ] 
reg.undes.Pitch.bRpR <- reg.undesirable[reg.undesirable$PBhand == "bRpR", ]   

reg.undes.Pitch.bRpR.count<-table(reg.undes.Pitch.bRpR$pitch_cat, reg.undes.Pitch.bRpR$count)

reg.undes.Pitch.bRpR.count.prop<-round(prop.table(reg.undes.Pitch.bRpR.count,2)*100,digits=2)
```

```{r}
#creation of undesirable bRpR matrix

reg.undes.Pitch.bRpR.count.matrix<-as.matrix(reg.undes.Pitch.bRpR.count.prop)

reg.undes.pitch.bRpR.count.prop<-reg.undes.Pitch.bRpR.count.matrix %*% UD_RHP_Prop_I_matrix

colnames(reg.undes.pitch.bRpR.count.prop) <- c('0-0', '0-1', '0-2', '1-0', '1-1', '1-2', '2-0', '2-1', '2-2', '3-0', '3-1', '3-2')
reg.undes.pitch.bRpR.count.prop
```



Probabilities for bRpR 
```{r}
#change name of variable; proportions for desirable pitches thrown for bRpR
prop.matrix.D.bRpR<-reg.des.pitch.bRpR.count.prop
#change name of variable; proportions for undesirable pitches thrown for bRpR
prop.matrix.U.bRpR<-reg.undes.pitch.bRpR.count.prop

#normalize the matrices
#creation of the probabilities; probability of a desirable outcome occuring given it is a bRpR instance

normal.D.matrix.bRpR<-prop.matrix.D.bRpR/(prop.matrix.D.bRpR + prop.matrix.U.bRpR)
normal.D.matrix.bRpR
```
These are the probabilities that will be tested using a Chi Square Test to see if the distributions of the probabilites are statistically different from a uniform distribution (average probability of the three probabilities (each pitch_cat) by count)



Statistical tests
_____________________________________________
_____________________________________________


Chi Squared Test


#lines 293-323 are for readers understanding of n-value breakdown; not used in actual chi squared test code

```{r}
#n-values for desirable bRpR

chi_D_bRpR<-reg.outcome %>% 
  filter(p_outcome== "desirable" & PBhand == "bRpR" )%>%
  drop_na(pitch_cat,PBhand, count, p_outcome) %>%
  group_by(count,pitch_cat) %>%
  summarize(D_bRpR = n()) %>%
  group_by(count,pitch_cat)
chi_D_bRpR
```

```{r}
#n-values for undesirable bRpR 

chi_U_bRpR<-reg.outcome %>% 
  filter(p_outcome== "undesirable" & PBhand == "bRpR" )%>%
  drop_na(pitch_cat,PBhand, count, p_outcome) %>%
  group_by(count,pitch_cat) %>%
  summarize(U_bRpR = n()) %>%
  group_by(count,pitch_cat) 
chi_U_bRpR
```

```{r}
#take U_bRpR and add column to existing chi_D_bRpR, then change to new dataframe name

chi_D_bRpR$U_bRpR <- chi_U_bRpR$U_bRpR
chi_bRpR_totals<-chi_D_bRpR
chi_bRpR_totals
```


```{r}

#code to be used in chi squared test on data

chi_bRpR<-reg.outcome %>% 
  filter(PBhand == "bRpR" )%>%
  drop_na(pitch_cat,PBhand, count, p_outcome) %>%
  group_by(count,pitch_cat)
#select rows
chi_bRpR<-chi_bRpR[,c(2,8,10)]
chi_bRpR
```

Now for the chi-square test; need to make a contingency table

```{r}
#0-0 count
#filter dataframe above for 0_0 count
chi_bRpR_0_0<-chi_bRpR%>% 
  filter(count=="0_0")

table(chi_bRpR_0_0$pitch_cat, chi_bRpR_0_0$p_outcome)

#code for test; test passes n-values shown in table above for determination of p-value
chisq.test(chi_bRpR_0_0$pitch_cat, chi_bRpR_0_0$p_outcome)
```
#p-value shows that there is a statistically significant difference in the distributions of the probabilites by count compared to a uniform distributioun. 
#p-value needed to be less than 0.05 to be considered significantly different

```{r}
#0-1 count
chi_bRpR_0_1<-chi_bRpR%>% 
  filter(count=="0_1")

table(chi_bRpR_0_1$pitch_cat, chi_bRpR_0_1$p_outcome)
chisq.test(chi_bRpR_0_1$pitch_cat, chi_bRpR_0_1$p_outcome)
```

#The other counts are not shown as they are the same; copy and paste and change filter(count=="0_1") to be count of interest
#all p-values for each of the 12 counts were used in the visual below

___________________________________________

Pairwise Comparison of Proportion Test

statistical test: Pairwise comparison for proportions
Pairwise comparison means comparing all pairs of something. If I have three items A, B and C, that means comparing A to B, A to C, and B to C.

In this case the test will be used to see if there is a statistically significant difference between fastball, breaking, offspeed probabilities within each count


```{r}

pitches_bRpR<-reg.outcome %>% 
  filter(PBhand == "bRpR" )%>%
  drop_na(pitch_cat, PBhand, count, p_outcome) %>%
  group_by(count, pitch_cat)

pitches_bRpR = subset(pitches_bRpR, select = c("count", "pitch_cat", "p_outcome"))
pitches_bRpR
```
I will show how I got results for both the 0-0 and 0-1 counts. Again, the other counts can be found by just changing the count you would like to see

```{r}

 # Creates a new numeric variable that is p_outcome as a 0 or 1
pitches_bRpR$p_outcome.n = as.numeric(pitches_bRpR$p_outcome) - 1 
pitches_bRpR_0_0 <-pitches_bRpR %>% filter(count=="0_0")
```

```{r}
#used for pairwise test
pitchTab_bRpR_0_0<-xtabs( ~ pitch_cat + p_outcome, 
       data=pitches_bRpR_0_0)
round(prop.table(pitchTab_bRpR_0_0, margin = 1),5)
```
```{r}
pairwise.prop.test(x = pitchTab_bRpR_0_0[,c("desirable","undesirable")])
```

Based on the results above, the F-B pair shows evidence of a significant difference between the probabilities (p < 0.05).

I will now change the count to be 0-1
```{r}

 # Creates a new numeric variable that is p_outcome as a 0 or 1
pitches_bRpR$p_outcome.n = as.numeric(pitches_bRpR$p_outcome) - 1 
pitches_bRpR_0_1 <- pitches_bRpR %>% filter(count=="0_1")

```

```{r}

pitchTab_bRpR_0_1<-xtabs( ~ pitch_cat + p_outcome, 
       data=pitches_bRpR_0_1)

round(prop.table(pitchTab_bRpR_0_1, margin = 1),5)
```
```{r}
pairwise.prop.test(x = pitchTab_bRpR_0_1[,c("desirable","undesirable")])
```

Looking at the results for the 0_1 count, the F-B pair shows evidence of a significant difference between the probabilities. The other counts were done before creating the visual below. 


____________________________________________________________________________________
____________________________________________________________________________________

ggplot visual

I will now show how I created the visuals that are seen in my write up
Below is the code I used to do so. I understand that there is probably a more consise way to do it however this is how I made it work.

```{r}
#dataframe taken from line 276; converts matrix to dataframe
bRpR_D_matrix<-as.data.frame(normal.D.matrix.bRpR)
bRpR_D_matrix

#transpose matrix moving columns(count) to rows
bRpR_D_T_matrix<-t(bRpR_D_matrix)

#convert matrix to dataframe
bRpR_D_df<-as.data.frame(bRpR_D_T_matrix)

#make row names to its own column (count)
bRpR_D_df <- data.frame(count = row.names(bRpR_D_df), bRpR_D_df, row.names=NULL)
bRpR_D_df<-bRpR_D_df
bRpR_D_df
```

```{r}
#convert to data frame
total.reg.bRpR.df<-as.data.frame(total.reg.bRpR)
total.reg.bRpR.df
```

```{r}
D_bRpR<-reg.outcome %>% 
  filter(p_outcome== "desirable" & PBhand == "bRpR" )%>%
  drop_na(pitch_cat,PBhand, count, p_outcome) %>%
  group_by(count,pitch_cat) %>%
  summarize(D_bRpR = n()) %>%
  group_by(count,pitch_cat)
D_bRpR
```


```{r}
#library(tidyr)

#move each of the pitch type columns to row factors
bRpR_D_df.long<- pivot_longer(bRpR_D_df, cols=2:4, names_to = "Pitch_Type", values_to = "Prob_Desirable")
bRpR_D_df.long


#arrange pitch_type rows by factor
bRpR_D_df2<- bRpR_D_df.long %>% arrange(count)
bRpR_D_df2<-cbind(bRpR_D_df2, Total_bRpR=total.reg.bRpR.df$Freq, Total_D_bRpR = D_bRpR$D_bRpR)
bRpR_D_df2
```

```{r}
# 'bRpR_signif_vec' is used from the pairwise test results, needed for the asterick placed above certain bars in visual below
#0 - no, 1 - yes
bRpR_signif_vec<-c(0,1,0,0,1,0,0,1,1,0,0,0,0,0,0,0,1,0,0,0,0,0,0,0,0,1,0,0,0,0,0,0,1,0,0,1)
bRpR_signif_vec<- as.factor(bRpR_signif_vec)
bRpR_D_df2$Significant <-bRpR_signif_vec
bRpR_D_df2
```
Using the dataframe above, the visual below was created. 

```{r, fig.width=4}

plot_bRpR <- ggplot(bRpR_D_df2, aes(x = count, y = Prob_Desirable, fill = Pitch_Type)) +
  geom_col(width=0.78,position = position_dodge(width=0.76),size=0.70, colour="black" ) +
  scale_fill_brewer(palette="Paired")+
  #below is how the "Significant" column wasn used to place astericks in the visual above respective bars
  geom_text(aes(label = ifelse(Significant == "1", "*", ""), group = Pitch_Type), 
            position = position_dodge(width = .9), vjust = 0.1, size = 20 / .pt, color="red")+
  theme_light()+
  labs(x = "Count, Total Pitches, Chi-Squared Results",
       y="Probability of Desirable Pitch",
       fill = "Pitch Type")+
  theme(axis.text.x = element_text(hjust=0.5, size=9), axis.text = element_text(face="bold"), 
        legend.position = c(.15, .98),
    legend.justification = c("right", "top"),
    legend.box.just = "right",
    legend.margin = margin(6, 6, 6, 6))+
  ggtitle("Pitch Type by Count (Desirable Outcomes)", "Regular season pitches for bRpR ")+
  ylim(c(0,1))

#my own x-axis was created to accommodate the count, n-values, and chi squared test results. Results that are not shown for the chi-squared test are results that had a p-value less than 0.05. 

plot_bRpR <- plot_bRpR + scale_x_discrete("Count, n-value, χ² Result", labels = c("0-0" = "0-0\nn=39155\n χ²=            ", "0-1" = "0-1\n18973", "0-2" = "0-2 \n8820", "1-0" = "1-0 \n15434\n0.425", "1-1" = "1-1 \n 15396\n0.072", "1-2" = "1-2 \n13734", "2-0" = "2-0 \n5301\n0.825", "2-1" = "2-1 \n8328\n0.834", "2-2" = "2-2 \n12933", "3-0" = "3-0 \n1641\n0.906", "3-1" = "3-1 \n3420", "3-2" = "3-2 \n8184"))
plot_bRpR
```



This is the code I used for the bRpR category in the regular season. I used the same code for the other pitcher/batter handedness matchups however it is not shown to shorten the length of this notebook.

Please review my formal write up of my project for more of an explanation on the process and statistic results. 

You can view my project at https://amandapoor.net

